### 2019-07-01 submission/manuscript.tex
Revisions and addressing Nick's comments

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
Mice were treated with the PPI omeprazole for 7 days prior to \emph{C.
difficile} challenge and were compared to mice that were treated with
the antibiotic clindamycin or both omeprazole and clindamycin. We found
that omeprazole was not sufficient to promote \emph{C. difficile}
colonization. When omeprazole treatment was combined with the
antibiotic, there was a mixed level of \emph{C. difficile} colonization,
where one cage of mice remained resistant and the other cage was
colonized. 16S rRNA sequencing analysis revealed that omeprazole had
minimal impact on the murine microbiota throughout the 16 days of PPI
exposure. \emph{C. difficile} colonization of clindamycin-treated mice
was associated with changes in \emph{Alistipes}, \emph{Barnesiella},
Porphyromonadaceae, and Ruminococcaceae. These results suggest PPI
treatment alone is not sufficient to disrupt microbiota resistance to C.
difficle infection in mice that are normally resistant in the absence of
antibiotic treatment.

@@

Antibiotics are a major risk factor for \emph{Clostridium difficile}
infections (CDIs), but it is less clear what other factors contribute to
CDI incidence and recurrence. Interestingly, other medications besides
antibiotics have been shown to alter the microbiota. Proton pump
inhibitors (PPIs) are associated with alterations in the human
intestinal microbiota through observational and interventional stduies
and PPI use has also been associated with CDIs. We evaluated the ability
of the PPI omeprazole to alter the murine intestinal microbiota and
disrupt colonization resistance to \emph{C. difficile}. We found PPI
treatment had minimal impact on the murine intestinal microbiota and was
not enough to promote \emph{C. difficile} colonization after challenge.
Further studies are needed to determine whether other factors such as
the composition of the starting bacterial community, comorbidities, and
use of additional medications contribute to the association between PPIs
and CDIs.

@@

Proton pump inhibitors (PPIs) are one of the most widely used drugs and
are available over the counter (1). PPIs are approved to treat a variety
of conditions including gastroesophageal reflux disease, ulcers, or as
prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated
gastrointestinal bleeding (2). Unfortunately, there have been a number
of adverse events associated with their use including small intestinal
bacterial overgrowth, decreased calcium absorption, and increased risk
of enteric infections (2). There's also been a number of large cohort
studies as well as interventional clinical trials that have demonstrated
specific alterations in the intestinal microbiome were associated with
PPI use (3, 4). PPI-associated microbiota changes have been attributed
to the ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (3, 4). Human fecal microbiota
changes with PPI use include increases in Enterococcaceae,
Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
with decreases in Ruminococcaceae (3). An \emph{in vitro screen},
demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
Lachnospiraceae and Ruminococcaceae, which represented
\textasciitilde{}16\% of the 40 intestinal bacteria strains that were
screened (5).\\
Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (6). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between PPI use and incidence or recurrence of CDIs (2,
4, 7, 8). However, most of the studies suggesting a link between PPIs
and \emph{C. difficile} are retrospective and did not evaluate the
microbiome (2, 7, 8). Thus, it is unclear whether the gastrointestinal
microbiome changes associated with PPI use play a role in the
association between PPIs and CDI incidence or recurrence. Additionally,
epidemiological studies have a limited capacity to address potential
confounders and comorbidities in patients that were on PPIs and
developed CDIs or recurrent CDIs (7, 8).\\
Although there have been a few \emph{C. difficile} mouse model studies
that have demonstrated PPIs have some effect on CDIs with or without
additional antibiotic treatment, none of these studies looked at the
interplay with the intestinal microbiome (9--11). Here, we evaluated the
impact of a PPI on the murine microbiome and susceptibility to \emph{C.
difficile} colonization in relation to clindamycin, an antibiotic that
perturbs the microbiome enough to allow \emph{C. difficile} to colonize
but is mild enough that \emph{C. difficile} is cleared within 10 days
(12).

@@

\textbf{Murine fecal microbiomes were minimally affected by PPI
treatment } To test if PPI treatment alters the microbiome and promotes
susceptibility to CDIs, we gavaged mice daily with a high dose of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A)
and compared the PPI-treated mice to a group that received the
antibiotic, clindamycin, and a group that received both PPI and
clindamycin treatment. A principle coordinates analysis (PCoA) of the
Bray-Curtis distances over the initial 7 days of treatment revealed the
bacterial communities of PPI-treated mice remained relatively unchanged
(Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
that have previously been impacted by PPIs in human studies, but we see
no significant fluctuations in the stool of our PPI-treated mice
throughout the course of the 16-day experiment (Figure 1C). In contrast,
the microbiomes from the 2 groups of mice that received clindamycin
started to shift away from the untreated and PPI-treated mice the day
after treatment (Figure 1D).

@@

Our results demonstrate the PPI omeprazole had minimal impact on the
stool microbiomes of mice and did not promote susceptibility to \emph{C.
difficile} colonization. In contrast, clindamycin-treated mice were
initially colonized, but cleared the infection within 10 days. Combining
PPI and clindamycin treatment had a mixed effect, one cage was colonized
with \emph{C. difficile} while the other cage remained resistant.\\

@@

taxa observed to be significantly impacted by PPI use in human studies,
but a closer examination suggests that the significance of these
associations in humans is driven by some individuals more than others
and the effect size can be relatively small with overall differences on
a PCoA plot difficult to distinguish (14). Also, most of the studies to

@@

Previous work that administered the PPI lansoprazole daily for 2 weeks
to mice and then challenged with \emph{C. difficile} demonstrated that
PPI treatment alone resulted in detectable \emph{C. difficile} in the
stool 1 week after challenge, but also showed there was detectable
\emph{C. difficile} in mice not treated with antibiotics (10, 11). The
presence of \emph{C. difficile} in mice that were not treated with
either antibiotics or PPIs could be attributed to the higher dose of
vegetative cells (10\textsuperscript{8} CFU) and the different \emph{C.
difficile} strain used to challenge the mice and may partly explain
their observation of \emph{C. difficile} in PPI-treated mice (10, 11).
We have previously shown mice from are colony that are not given
antibiotics are resistant to \emph{C. difficile} 630 when challenged
with 10\textsuperscript{3} spores (17). Another group also used a higher
dose of 10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative
cells and a different \emph{C. difficile} strain to challenge
antibiotic-treated mice or mice treated with both the PPI esomeprazole
for 2 days and antibiotics and demonstrated the antibiotic/PPI-treated
mice developed more severe CDIs (9).\\

@@

\emph{Streptococcus} and \emph{Enterococcus} are rare taxa in our
C57BL/6 mouse colony. If these taxa are important contributors to the

@@

between PPIs and CDIs. NSAID treatment can increase CDI severity in mice
(18). Given that PPIs are sometimes used in combination with NSAIDs (2),
it may be worth examining both medications together using a mouse model.
This study only looked at the impact of PPIs with or without
antibiotics, but future work may want to incorporate age, other
comorbidities and bacterial strains that are less common in mice to
evaluate whether PPIs in conjunction with other factors can increase the
risk of CDIs.

@@

\textbf{Drug treatments} Omeprazole (Sigma Aldrich) was prepared in a
vehicle solution of 40\% polyethylene glycol 400 (Sigma-Aldrich) in
phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
aliquots and diluted to an 8mg/mL prior to gavage. All mice received
\textasciitilde{}40 mg/kg omeprazole or vehicle solution once per day
through the duration of the experiment with treatment starting 7 days
before \emph{C. difficile} challenge (Figure 1A). One day prior to
\emph{C. difficile} challenge, mice received an intraperitoneal
injection of 10 mg/kg clindamycin or sterile saline vehicle. All drugs
were filter sterilized through a 0.22 micron syringe filter before
administration to animals.

\textbf{C. difficile infection model} Mice were challenged with \emph{C.
difficile} 630 seven days after the start of omeprazole treatment and
one day after clindamycin treatment. Mice were challenged with
10\textsuperscript{3} spores in ultrapure distilled water. Samples were
collected for 16S rRNA sequencing or \emph{C. difficile} CFU
quantification throughout the duration of the experiments at the
indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were
flash frozen in liquid nitrogen and stored at -80°C until DNA
extraction, while samples for CFU quantification were transferred into
an anaerobic chamber and serially diluted in PBS. Diluted samples were
plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar)
plates and incubated at 37°C for 24 hours under anaerobic conditions to
quantify \emph{C. difficile} CFU.

@@

\textbf{Statistical Analysis}

@@

\textbf{Figure 1. PPI treatment had minimal impact on the murine fecal
microbiota} A. Mouse experiment timeline and logistics. B. Principal
Coordinates Analaysis (PCoA) of Bray-curtis distances from stool samples
for all treatment groups during the intial 7 days of treatment with the
PPI omeprazole or vehicle cluster together and do not chage much over
time. C. Relative abundances of families previously associated with PPI
use in humans do not change much over the 16-day course of treatment
with PPIs (7 days before \emph{C. difficile} challenge through 9 days
post \emph{C. difficile} challenge). D. PCoA of stool samples collected
before and 1 day after antibiotic treatment, demonstrating the fecal
microbiota of mice treated with clindamycin starts to separate from the
rest of the samples.

\textbf{Figure 2. PPI treatment alone does not promote CDIs in mice} A.
C. difficile CFUs/g stool measured each day post C. difficile challenge
for clindamycin, clindamycin/PPI, and PPI-treated mice. B. PCoA of stool
samples collected after antibiotic treatments. C. Genera significantly
associated with treatment groups. D. Families significantly associated
with treatment groups.

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

@@

>>>>

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether PPIs promote CDIs using a CDI mouse model.
Mice were treated with a high daily dose of the PPI omeprazole for 7
days prior to \emph{C. difficile} challenge and were compared to mice
that were treated with the antibiotic clindamycin or both omeprazole and
clindamycin. High-dose PPI treatment was maintained throughout the
experiment, which included \emph{C. difficile} challenge and the
following 9 days. We found that omeprazole was not sufficient to promote
\emph{C. difficile} colonization. When omeprazole treatment was combined
with the antibiotic, there was a mixed level of \emph{C. difficile}
colonization, where one cage of mice remained resistant and the other
cage was colonized. 16S rRNA sequencing analysis revealed that
omeprazole had minimal impact on the murine microbiota throughout the 16
days of PPI exposure. \emph{C. difficile} colonization of
clindamycin-treated mice was associated with changes in
\emph{Alistipes}, \emph{Barnesiella}, Porphyromonadaceae, and
Ruminococcaceae. These results suggest PPI treatment alone is not
sufficient to disrupt microbiota resistance to C. difficle infection in
mice that are normally resistant in the absence of antibiotic treatment.

@@

Antibiotics are a major risk factor for \emph{Clostridium difficile}
infections (CDIs), but it is less clear what other factors contribute to
CDI incidence and recurrence. Interestingly, other medications besides
antibiotics have been shown to alter the microbiota. Proton pump
inhibitors (PPIs) are associated with alterations in the human
intestinal microbiota through observational and interventional studies
and PPI use has also been associated with CDI incidence and recurrence.
We evaluated the ability of a high dose of the PPI omeprazole to alter
the murine intestinal microbiota and disrupt colonization resistance to
\emph{C. difficile}. We found daily PPI treatment had minimal impact on
the murine intestinal microbiota and was not enough to promote \emph{C.
difficile} colonization after challenge. Further studies are needed to
determine whether other factors such as the composition of the starting
bacterial community, comorbidities, and use of additional medications
contribute to the association between PPIs and CDIs.

@@

Proton pump inhibitors (PPIs) are one of the most widely used drugs and
are available over the counter (1). PPIs are approved to treat a variety
of conditions including gastroesophageal reflux disease, ulcers, or as
prophylaxis for nonsteroidal anti-inflammatory drug (NSAID)-associated
gastrointestinal bleeding (2). Unfortunately, there have been a number
of adverse events associated with their use including small intestinal
bacterial overgrowth, decreased calcium absorption, and increased risk
of enteric infections (2). There's also been a number of large cohort
studies as well as interventional clinical trials that have demonstrated
specific alterations in the intestinal microbiome were associated with
PPI use (3, 4). PPI-associated microbiota changes have been attributed
to the ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (3, 4). Human fecal microbiota
changes with PPI use include increases in Enterococcaceae,
Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
with decreases in Ruminococcaceae (3). An \emph{in vitro screen},
demonstrated that 3 PPIs inhibit the growth of 7-8 strains of
Lachnospiraceae and Ruminococcaceae, which represented
\textasciitilde{}16\% of the 40 intestinal bacteria strains that were
screened (5).\\
Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (6). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between PPI use and incidence or recurrence of CDIs (2,
4, 7, 8). However, most of the studies suggesting a link between PPIs
and \emph{C. difficile} are retrospective and did not evaluate the
microbiome (2, 7, 8). Thus, it is unclear whether the gastrointestinal
microbiome changes associated with PPI use play a role in the
association between PPIs and CDI incidence or recurrence. Additionally,
epidemiological studies have a limited capacity to address potential
confounders and comorbidities in patients that were on PPIs and
developed CDIs or recurrent CDIs (7, 8).\\
Although there have been a few \emph{C. difficile} mouse model studies
that have demonstrated PPIs have some effect on CDIs with or without
additional antibiotic treatment, none of these studies looked at the
interplay with the intestinal microbiome (9--11). Here, we evaluated the
impact of a daily high dose PPI treatment on the murine microbiome and
susceptibility to \emph{C. difficile} colonization in relation to
clindamycin, an antibiotic that perturbs the microbiome enough to allow
\emph{C. difficile} to colonize but is mild enough that \emph{C.
difficile} is cleared within 10 days (12).

@@

\textbf{Murine fecal microbiomes were minimally affected by PPI
treatment } To test if PPI treatment alters the microbiome and promotes
susceptibility to CDIs, we gavaged mice daily with a high dose of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A)
and compared the PPI-treated mice to a group that received the
antibiotic, clindamycin, and a group that received both PPI and
clindamycin treatment. A principle coordinates analysis (PCoA) of the
Bray-Curtis distances over the initial 7 days of treatment revealed the
bacterial communities of PPI-treated mice remained relatively unchanged
(Figure 1B). Enterococcaceae, Lactobacillaceae, Micrococcaceae,
Staphylococcaceae, Streptococcaceae, and Ruminococcaceae are all familes
that have previously been impacted by PPIs in human studies, but we see
no significant fluctuations in the stool of our PPI-treated mice
throughout the course of the 16-day experiment (Figure 1C, Figure S1).
In contrast, the microbiomes from the 2 groups of mice that received
clindamycin started to shift away from the PPI-treated mice and the
samples from previous days one day after treatment (Figure 1D).

@@

Our results demonstrate that daily treatment with a high dose of the PPI
omeprazole had minimal impact on the stool microbiomes of mice and did
not promote susceptibility to \emph{C. difficile} colonization. In
contrast, clindamycin-treated mice were initially colonized, but cleared
the infection within 10 days. Combining PPI and clindamycin treatment
had a mixed effect, one cage was colonized with \emph{C. difficile} and
cleared within 10 days, while the other cage remained resistant.\\

@@

taxa observed to be significantly impacted by PPI use in human studies.
However, 3 of the families that typically increase were absent or at low
abundance in the mice from the beginning and may be one contributing
factor to why we did not see any changes. Additionally, some of the
significance of PPI associations in human interventional trials appears
to be driven by a handful of specific taxa with overall differences on a
PCoA plot difficult to distinguish (14). Also, most of the studies to

@@

Previous work that administered 0.5 mg/kg of the PPI lansoprazole daily
for 2 weeks to mice and then challenged with \emph{C. difficile}
demonstrated that PPI treatment alone resulted in detectable \emph{C.
difficile} in the stool 1 week after challenge, but also showed there
was detectable \emph{C. difficile} in mice not treated with antibiotics
(10, 11). The presence of \emph{C. difficile} in mice that were not
treated with either antibiotics or PPIs could be attributed to the
higher dose of vegetative cells (10\textsuperscript{8} CFU) used to
challenge the mice and may partly explain their observation of \emph{C.
difficile} in PPI-treated mice (10, 11). We have previously shown mice
from our colony that are not given antibiotics are resistant to \emph{C.
difficile} 630 when challenged with 10\textsuperscript{3} spores (17).
Another group also used a higher dose of 10\textsuperscript{6} spores or
10\textsuperscript{7} vegetative cells to challenge antibiotic-treated
mice or mice treated with both the PPI esomeprazole (40mg/kg dose) for 2
days and antibiotics and demonstrated the antibiotic/PPI-treated mice
developed more severe CDIs (9).\\

@@

\emph{Streptococcus} and \emph{Enterococcus} are rare genera in our
C57BL/6 mouse colony. If these genera are important contributors to the

@@

between PPIs and CDI incidence or recurrence. NSAID treatment can
increase CDI severity in mice (18). Given that PPIs are sometimes used
in combination with NSAIDs (2), it may be worth examining both
medications together using a mouse model. This study only looked at the
impact of PPIs with or without antibiotics, but future work may want to
incorporate age, other comorbidities and bacterial strains that are less
common in mice to evaluate whether PPIs in conjunction with other
factors can increase the risk of CDIs or recurrent CDIs.

\subsection{Acknowledgements}\label{acknowledgements}

This research was supported by NIH grant U01AI12455. We would also like
to thank the Unit for Laboratory Animal Medicine at the University of
Michigan for maintaining our mouse colony and providing the
infrastructure and support for performing our mouse expeirments. The
authors are also thankful to members of the Schloss lab for helpful
discussions throughout the process of designing the experiment,
analyzing the results and drafting of the manuscript.

@@

\textbf{Drug treatments} Omeprazole (Sigma Aldrich) was prepared in a
vehicle solution of 40\% polyethylene glycol 400 (Sigma-Aldrich) in
phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
aliquots and diluted to an 8mg/mL prior to gavage. All mice received 40
mg/kg omeprazole or vehicle solution once per day through the duration
of the experiment with treatment starting 7 days before \emph{C.
difficile} challenge (Figure 1A). One day prior to \emph{C. difficile}
challenge, 2 groups of mice received an intraperitoneal injection of 10
mg/kg clindamycin or sterile saline vehicle. All drugs were filter
sterilized through a 0.22 micron syringe filter before administration to
animals.

\textbf{\emph{C. difficile} infection model} Mice were challenged with
\emph{C. difficile} 630 seven days after the start of omeprazole
treatment and one day after clindamycin treatment. Mice were challenged
with 10\textsuperscript{3} spores in ultrapure distilled water. Stool
samples were collected for 16S rRNA sequencing or \emph{C. difficile}
CFU quantification throughout the duration of the experiments at the
indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were
flash frozen in liquid nitrogen and stored at -80°C until DNA
extraction, while samples for CFU quantification were transferred into
an anaerobic chamber and serially diluted in PBS. Diluted samples were
plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar)
plates and incubated at 37°C for 24 hours under anaerobic conditions to
quantify \emph{C. difficile} CFU.

@@

\textbf{Statistical Analysis} To assess for differences in relative
abundance in families and genera across our 3 different treatment groups
(Clindamycin, Clindamycin + PPI, and PPI), we used a Kruskal-Wallis test
with a Benjamini-Hochberg correction for multiple comparisons.

@@

\subsection{References}\label{references}

@@

\textbf{Figure 1. PPI treatment had minimal impact on the murine fecal
microbiota} A. Mouse experiment timeline and logistics. Stools for 16S
rRNA sequencing analysis were collected on the days that are labeled
(Day -7, -5, -3, -1, 0, 1, 2, 3, 4, 5, 7, 9). B. Principal Coordinates
Analaysis (PCoA) of Bray-curtis distances from stool samples for all
treatment groups during the intial 7 days of treatment with the PPI
omeprazole or vehicle cluster together and do not chage much over time.
C. Relative abundances of families previously associated with PPI use in
humans do not change much over the 16-day course of treatment with PPIs
(7 days before \emph{C. difficile} challenge through 9 days post
\emph{C. difficile} challenge). D. PCoA of stool samples collected
before and 1 day after antibiotic treatment, demonstrating the fecal
microbiota of mice treated with clindamycin starts to separate from the
rest of the samples.

\textbf{Figure 2. PPI treatment alone does not promote CDIs in mice} A.
C. difficile CFUs/g stool measured each day post C. difficile challenge
for clindamycin, clindamycin/PPI, and PPI-treated mice. Lines represent
the mean CFU/g for each treatment group while points represent CFU/g for
individual mice within each group. B. PCoA of stool samples collected
after antibiotic treatments. C. Genera significantly associated with
treatment groups for samples across all sequenced timepoints. D.
Families significantly associated with treatment groups for samples
across all sequenced timepoints. C,D data were analyzed by
Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
comparisons.

\textbf{Figure S1. Previous human PPI-associated families fluctuate over
time with no overall trend in either direction} Relative abundance over
time for Lactobacillaceae (A) and Ruminococcaceae (B), 2 of the
PPI-associated familes from human PPI studies across all 3 treatment
groups. Each point represents the relative abundance for an individual
mouse
